---
title: "Untitled"
author: "Michael Ching, MD, MPH"
date: "December 16, 2016"
output: html_document
---

I sit at a Pokestop at work and I've been gathering data on what comes out each time I spin the disk. The questions that I had were: (1) are there differences between Pokestops in what you get? (2) do Pokestops change what they give over time? Here's some analysis looking at those questions.

The dataset is available for viewing [here](https://docs.google.com/spreadsheets/d/1FmApMo7dwv9JI6qSVKWKGv9iQU23isrbbbcvn0TVsZw/edit?usp=sharing).
```{r}
library(dplyr)
x <- read.csv("../datasets/pokestops.csv")
x <- tbl_df(x)
x <- x[-13] # remove Total column since this will make calculations easier

# Format Date variable
x$Date <- as.character(x$Date)
x$Date <- paste0(x$Date, "/2016")
x$Date <- as.Date(x$Date, "%m/%d/%Y")
```

Because of how I entered the data, all the blank spaces were coerced into NAs. I had to process these into 0s.
```{r}
NAto0 <- function(z) {
  z[is.na(z)] <- 0
  return(z)
}

for(v in 1:length(x)) {
  x[v] <- NAto0(x[v])
}
```

There were `r nrow(x)` spins with a total of `r sum(x[,-c(1:3)])` items received. 


Here are the mean number of items received per throw. I would disregard the stats on Max Potions since I reached level 25 at the very end of the analysis.
```{r}
sapply(x[-c(1:3)], mean)
```

Here is the overall frequency of what the Pokestops gave and the percentage breakdown.
```{r}
item_totals <- summarize_each(x[-c(1:3)], funs(sum))
item_totals
item_totals/sum(item_totals)
```

```{r}
x_by_date <- x %>% select(-c(2:3)) %>% 
  group_by(Date) %>% 
  select(-1) %>%
  summarize_each(funs(sum)) %>% 
  ungroup()
x_by_date
p_by_date <- prop.table(as.matrix(x_by_date[,-1]), 1)
p_by_date <- data.frame(p_by_date)
p_by_date <- bind_cols(data.frame(spin_date = x_by_date$Date), p_by_date)

```

```{r}
library(ggplot2)
library(tidyr)

item_names <- names(p_by_date)[-1] 
p_by_date <- p_by_date %>% gather("item", "p", 2:10)
p_by_date$item <- factor(p_by_date$item, levels = item_names)

ggplot(p_by_date, aes(x = spin_date, y = p, color = item)) + geom_point() +
  scale_color_brewer(palette = "Set3") + 
  ggtitle("Daily Variation in Item Drop Rate")
```

