---
title: "Work for Dinosaur Speed Geocache"
author: "Michael Ching"
date: "August 5, 2016"
output:
  html_document:
    keep_md: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

This is the work for the geocache we did today. The geocache description instructed us to measure the stride length and footprint length of some dinosaur tracks at the Dinosaur Ridge hike outside of Denver, Colorado. The kids and I looked at both the thinner Ornithomimus tracks and the thicker Iguanodontid tracks. Here's the work, done in R:

First, set up a table of the data. The units for stride length and footprint length are meters.
```{r}
dino_data <- data.frame(species = c("Ornithomimus", "Iguanodontid"),
                        stride_length = c(1.041, 1.063),
                        footprint = c(0.18, 0.415))

dino_data
```

Second, create an estimate of the height.

```{r}
dino_data$height <- dino_data$footprint * 4
dino_data
```

Third, create the velocity estimates in meters per second.

```{r}
g <- 9.81 # meters per second
dino_data$mps <- 0.25 * (g ^ 0.5) * 
  dino_data$stride_length^1.67 * 
  dino_data$height^(-1.17)
dino_data
```

Finally, convert to miles per hour by multiplying meters per second by 3600 seconds per hour and 1/1609 miles per meter.

```{r}
dino_data$mph <- dino_data$mps * 3600 / 1609
```

Here are the results:
```{r}
dino_data
```

Thanks for a fun cache!