---
output: 
  html_document: 
    keep_md: yes
---

---
layout: post
title: Sorting photo directories with R and exiftool
---
We were fans of Google's Picasa photo organization software, but when it was discontinued, I decided that I would like to transfer all my digital photos onto an external drive and then into Adobe Lightroom. To do so, however, my "research" told me that it would be best to group them into catalogs based on year, mainly because I have on the order of 10<sup>5</sup> photos.

Right now I have my pictures organized into about 300-400 directories by event. I decided I wanted to organize all of my folders as directories within year directories. So all the 2016 folders would need to be sorted into a 2016 directory, all the 2015 folders would go into a 2015 directory, etc. I wanted it to look something like this:

```
-2016
--Mom's birthday
--Valentine's Day
--Big Island Trip
-2015
--Schofield Days
--Bellows Camp
--etc...
```

The problem was most of the folders are titled things like "Kindergarten Graduation" or "Anacapa". Some of them had the date in them (Maui 2012), or I could remember the date (Our Wedding). But for about 350 of them, there was no date in the folder title, and to figure out when these pictures were from, I would have to manually look at each directory, check the date from the EXIF data, and then sort those into the proper folder. Even if it took 1 minute per folder, that would take me 6 hours. So I decided I could spend 6 hours doing this manually or 6 hours learning some way to automate it. 

# Exiftool to the Rescue! Automating Metadata Extraction
A quick search online brought me to [`exiftool`](http://www.sno.phy.queensu.ca/~phil/exiftool/). This is a command line program that allows for reading and writing photo metadata. It has lots of awesome quotes on the webpage, and the one that sealed the deal was this one:

> "... it is total f***ing gibberish to me." - [Reddit Linux Questions](https://www.reddit.com/r/linuxquestions/comments/2yiked/i_want_to_batch_extract_the_exif_datetime_from_10/)

After some fooling around, I discovered that I could extract the date at the command line and send it to a .csv (in Mac OS X). Adding other EXIF attributes allows you to extract more data, but since I didn't really want anything else, I just pulled the filename and the DateTimeOriginal attributes.

```
> exiftool -DateTimeOriginal -S -s -csv ./*/ > all_photos_dates.csv
```

Sweet! Now that's something I can work with!

# Back to R: Cleaning and Formatting the Raw Data
First, I needed to read in the data.

```{r}
library(dplyr)
library(stringr)
library(tidyr)
photos <- read.csv(file = "~/Dropbox/Mike/photo_analysis/all_photo_dates.csv")
photos <- tbl_df(photos)
```
Here's what the first 10 rows look like.
```{r}
photos
```

# Reformatting the Directory and File Name Structure
We can see that the `exiftool` command read in the file names in the format `./directoryname/filename`. I need to split out that directory name and the component files. We can do it using the `separate` function from the `tidyr` package.

```{r}
photos <- photos %>% separate(SourceFile, c("dot", "dname", "fname"), sep = "/", remove = TRUE)
photos <- photos[,2:4] # drop the first column of dots
photos
```

# Formatting the Date Column
Ok, next thing we need to do is to take care of that date column and put it into a form that R can work with. I used POSIXct because POSIXlt caused problems when trying to add it to the data frame. This is because POSIXlt is a list, and POSIXct represents the number of seconds since the beginning of 1970.

```{r}
photos$DateTimeOriginal <- as.POSIXct(strptime(photos$DateTimeOriginal, format = "%Y:%m:%d %H:%M:%S"))
```

One of the problems with POSIXct is that it's not as easy to get the year out compared to POSIXlt. No problem, we can just temporarily convert to POSIXlt and add 1900 (the start date of POSIXlt years).

```{r}
photos <- photos %>% mutate(year = as.POSIXlt(DateTimeOriginal)$year + 1900)
photos
```

# Summarizing the Years in Each Folder
Now we want to get a summary of the years of the photos in each folder. I used the mode function from [here](http://stackoverflow.com/questions/2547402/is-there-a-built-in-function-for-finding-the-mode) to find out what was the most common year of the photos in each folder.

```{r}
Mode <- function(x) {
  ux <- unique(x)
  ux[which.max(tabulate(match(x, ux)))]
}

folder_years <- photos %>% group_by(dname) %>% summarize(year = Mode(year))
folder_years
```

Finally it'd be great if we could get a list of all the folders corresponding to each year. To do this, I created an empty list then populated it based on the most common year of the photos in each folder.

```{r}
folder_output <- list()

for(i in sort(unique(folder_years$year))) {
  x <- folder_years %>% filter(year == i)
  folder_output[as.character(i)][1] <- x[,1]
}
```

It worked great, and here's a sampling:

```{r}
folder_output[["2010"]]
```

# Putting the Lists by Year into a Convenient Format
To actually move the directories, I'd like to get a string of the directory names with quotes around the names. The quotes are needed because of the spaces in most of the names. For example "Dayton's 5th birthday" "Gus Ryan Phillip Visit" etc.

To do this I created a helper function that takes the contents of a character vector, put quotes around them, and then writes it out.

```{r}
format_dir_list <- function(x) {
  z <- NA
  for (i in 1:length(x)) {
    z[i] <- paste0("\"", x[i], "\"")
  }
  return(cat(z))
}
```

Here's the code to print out the lists of folder names. I didn't run it because again, there's soooo many folders. 
```{r, eval = TRUE}
for (i in 1:16) {
  cat("\n")
  print(names(folder_output)[i])
  cat("\n")
  format_dir_list(folder_output[[i]])
  cat("\n")
}
```

All I have to do now is to copy and paste each series of folder names into the OS X Terminal command `mv [pasted list of folders] [destination directory]` and I'll be done!

# Conclusion
That took more than 6 hours to figure out, but I had fun learning and only a little frustration. I know I could probably create some bash script to do this same thing, but I don't know bash hardly at all, and anyway, I wanted to figure out how to do it in R. Good thing there's no Comments on this blog. I would be really sad if someone posted a really tiny bash script that could do this same thing!

I think the lesson learned here is that if you have lots of photo folders, it's  best to have some kind of directory organization structure on the hard drive, rather than relying on Picasa or some other organization program to keep your folders in order. This way it will be easier to migrate to a new solution, and in theory, you know where things are even when you don't know exactly what the folder is called. If I were looking for a picture from the first day of my son's third grade year, I would know to look in 2014, and that would narrow the search down from all my folders to just the 2014 folders.