---
title: "Untitled"
author: "Michael Ching"
date: "June 9, 2017"
output: html_document
---

```{r}
library(dplyr)
library(ggplot2)
library(tidyr)
DL <- read.csv("../datasets/DL2010_2016.csv")
DL <- tbl_df(DL)
str(DL)
```

```{r}
DL$on_DL <- as.POSIXct(DL$on_DL)
DL$off_DL <- as.POSIXct(DL$off_DL)
```

```{r}
summary(DL$days)
```

Negative DL days? Let's look those up...
```{r}
which(DL$days < 0)
DL[which(DL$days < 0),]
```


Some googling showed what happened with these DL stints. Russell Branyan started 2010 on the DL with a herniated disk ([source](http://waitingfornextyear.com/2010/03/russell-branyan-to-start-season-on-dl/)) and came off the DL around April 20 ([source](https://www.youtube.com/watch?v=1xrkb5elJnY)). Milton Bradley ended this season on DL after being placed on it in July for knee injury. Hairston was injured in September with a fractured right tibia. I'll manually recode Branyan's off DL date to April 20 and the others to the end of the season.

```{r}
DL %>% filter(Season == 2010) %>% summarize(max(off_DL))
```

```{r}
DL$off_DL[which(DL$days <0)] <- as.POSIXct(c("2010-04-20", "2010-10-03", "2010-10-03"))
DL$days[which(DL$days <0)] <- as.integer(DL$off_DL[which(DL$days <0)] - DL$on_DL[which(DL$days <0)])
```

Here is the summary of injury by anatomical location. 
```{r}
DL %>% filter(censored_right == F) %>% group_by(Location_recode) %>% 
  filter(length(days) > 2) %>% summarize(mean(days), sd(days)) 
```

Here is the summary by 

```{r}
DL %>% filter(censored_right == F) %>% group_by(Injury_type_recode) %>% 
  filter(length(days) > 2) %>% summarize(mean(days), sd(days))
```

