---
title: "Bank Shot Stats"
output: html_notebook
---

There's a game at the Iolani Fair called Bank Shot. You have to roll a ball on a ramp into a hole and win a prize. Get it in the green hole and you win tickets, and if you get it in the red hole, you win a big inflatable prize. At one point we were giving out prizes like crazy and I wondered what was the rate of hitting the red and green holes.

Here's a schematic of the game. The whole board is tilted up at the far end so that the ball will always roll towards the holes.

```
_________________
|  /         \  |
|/             \|
|       |       |
|       |       |
|       |       |
|       |       |
|       |       |
|       |       |
|       |       |
|       |       |
|       |       |
| G   G |       |
|       |       |
|   R   |       |  R = red hole
|       |       |  G = green hole
| G   G |   o   |  o = ball
|_______________|

Player stands here
```

```{r}
library(tidyverse)
library(binom)
x <- read.csv("./datasets/bankshot.txt", sep = " ", colClasses = "character", header = F)
x <- tbl_df(x)
x$V2 <- toupper(x$V2)
y <- gather(x, "roll", "outcome", 1:2)
y$roll <- ifelse(y$roll == "V1", 1L, 2L)
```

```{r}
table(y$outcome)
```

```{r}
round(prop.table(table(y$outcome)), 3)
```

```{r}
table(roll=y$roll, outcome=y$outcome)
```

