---
title: "Pidgey Evolution: Effects on Combat Power and Hit Points (Updated)"
author: "Michael Ching"
date: "January 8, 2017"
output: 
  html_document:
    keep_md: true
---
I was evolving some Pokemon in Pokemon Go and wondered how much the Pokemon's attributes change after evolution. The attributes of interest were combat power (CP) and hit points (HP). I focused my analysis from Pidgeys because I had the most data points for this species.

I collected data from a couple of days' evolution, available for viewing [here](https://docs.google.com/spreadsheets/d/1kzYLqVn0-sA0ffGscNco9QkkC_0bce22S7pafiUs0uE/edit?usp=sharing). Both times I was using a Lucky Egg.

```{r, echo=FALSE, message=FALSE, warning=FALSE}
library(dplyr)
library(ggplot2)
```
```{r}
x <- read.csv("../datasets/evolution2.csv")
x <- tbl_df(x)
pidgeys <- filter(x, pokemon == "Pidgey")
pidgeys
```

## Question 1: What is the relationship between CP before and CP after evolution?

To explore what happened to CP before and after evolution, I plotted these on a graph.

```{r}
ggplot(pidgeys, aes(x = CP_pre, y = CP_post)) + geom_point(shape = 1) + 
  ggtitle("CP pre and post evolution")
```

The relationship was roughly linear. I modeled the relationship using simple linear regression. 

```{r}
pidgey_CP_model <- lm(CP_post ~ CP_pre, data = pidgeys)
summary(pidgey_CP_model)
```

Based on the data, the estimated multipler was 1.95 with a standard deviation of 0.009. The model explained roughly 99.9% of the variation in CP after evolution. Here is the model in equation form.

$$ CP_{post} = 1.925 \times CP_{pre} + 0.781 + \epsilon$$

## Question 2: How does evolution affect HP?

To view the relationship between pre and post HP, I plotted HP before and after evolution. 
```{r}
ggplot(pidgeys, aes(x = HP_pre, y = HP_post)) + geom_point(shape = 1) + 
  ggtitle("HP pre and post evolution")
```

There was also a roughly linear relationship but it did not appear to be as tightly linear as the CP before and after.

```{r}
pidgey_HP_model <- lm(HP_post ~ HP_pre, data = pidgeys)
summary(pidgey_HP_model)
```

From these observations it looks like the model is:

$$HP_{post} = 1.539 \times HP_{pre} -0.120 + \epsilon$$

This model explained >99% of the variance in HP after evolution. 

## Question 3: What is the relationship between CP and HP?
I plotted the relationship between CP and HP for Pidgeys.

```{r}
ggplot(pidgeys, aes(x = CP_pre, y = HP_pre)) + geom_point(shape = 1) + 
  ggtitle("Pidgey CP and HP Relationship")
```

This did not look like a linear relationship at all. I tried to fit several models. First we tried the linear one as a baseline. This was actually pretty good, although the diagnostics don't look great.

```{r}
pidgey_CP_HP_model <- lm(HP_pre ~ CP_pre, data = pidgeys)
summary(pidgey_CP_HP_model)
plot(pidgey_CP_HP_model)
```

I next tried a logarithmic model. The diagnostics on this weren't great either. The R squared was actually even worse than the linear model.
```{r}
pidgey_logCP_HP_model <- lm(HP_pre ~ log(CP_pre), data = pidgeys)
summary(pidgey_logCP_HP_model)
plot(pidgey_logCP_HP_model)
```

I then tried a square root model. The R squared was much better, and the diagnostics looked better.
```{r}
pidgey_sqrtCP_HP_model <- lm(HP_pre ~ I(CP_pre^(0.5)), data = pidgeys)
summary(pidgey_sqrtCP_HP_model)
plot(pidgey_sqrtCP_HP_model)
```

Here's the model in equation form:

$$HP = 3.04 \times \sqrt{CP} - 0.017 + \epsilon$$
Those numbers were so close to round numbers that we could say:

$$HP \approx 3 \times \sqrt{CP}$$

## Discussion
There appears to be a linear relationship between pre and post evolution CP and pre and post evolution HP. The relationship between CP and HP appears to be exponential. 

There are some other people who have done similar analyses on the relationship between pre and post evolution CP. There are even calculators that provide estimates of the post evolution CP. Here's [one example](https://pogotoolkit.com/#selectedPokemon=016&combatPower=100).

The original version of this paper had a negative intercept in the CP evolution model. This model's intercept is not significantly different from 0. The implication of the original negative intercept was that one might have a Pigeotto that had lower CP than the original Pidgey. With the updated model, this is no longer possible. 

I was not able to find other HP evolution analysis online. It appears that at least for Pidgeys, there is a simple linear relationship between pre and post evolution HP.

Here's a data set that could be used for cross validation: https://www.openintro.org/stat/data/?data=pokemon

