---
title: "Goodreads"
output: html_document
---

I just finished my 2017 Reading Challenge on Goodreads. My goal was to read 15 books this year. Poking around the site I discovered that I could export my data. I decided to have a look to see what my reading habits looked like.


## Dataset
```{r}
library(dplyr)
library(tidyr)
library(ggplot2)
library(lubridate)
library(hrbrthemes)
books <- read.csv("../datasets/goodreads.csv", colClasses = "character")
books <- tbl_df(books)
```
The data were arranged into variables like Author and Title but also Pages, Publication Date, Date Read, My Rating, Average Rating, etc. 

```{r}
str(books)
```

## Data Cleaning
Some boring data cleaning code...

```{r}
# Factor author names
books$Author <- factor(books$Author)

# Factor bookshelves
books$Exclusive.Shelf <- factor(books$Exclusive.Shelf)

# Numeric ratings
books$My.Rating <- as.numeric(books$My.Rating)
books$Average.Rating <- as.numeric(books$Average.Rating)

# Number of Pages
books$Number.of.Pages <- as.integer(books$Number.of.Pages)

# Years
books$Year.Published <- as.integer(books$Year.Published)
books$Original.Publication.Year <- as.integer(books$Original.Publication.Year)

# Dates
books$Date.Added <- ymd(books$Date.Added)
books$Date.Read <- ymd(books$Date.Read)
```

## Books Read vs Added
I've recorded 150 books on the website as being read, and have listed 16 books to read. There were 2 books that I'm sort of currently reading.

```{r}
books %>% select(Exclusive.Shelf) %>% 
  group_by(Exclusive.Shelf) %>% 
  summarize(n = length(Exclusive.Shelf))
```

## Dates Added and Read
I have only been adding to this list off and on since joining Goodreads. I plotted below the distribution of when I added and read books.

```{r}
tmp1 <- books %>% select(Book.Id, Date.Added) %>% mutate(action = "added") %>% rename(year = Date.Added)
tmp2 <- books %>% select(Book.Id, Date.Read) %>% mutate(action = "read") %>% rename(year = Date.Read)
bind_rows(tmp1, tmp2) %>% filter(!is.na(year)) %>%
  ggplot(aes(x = year, fill = action)) +
  geom_histogram(binwidth = 365, position=position_dodge()) +
  ggtitle("Books Added and Read per Year") +
  ylab("number of books") +
  xlab("year") +
  theme_ipsum()
```

It looks like I signed up for Goodreads in 2012 and started adding books to my list of read books. If I couldn't remember when I read the book, I left the date read field blank. After this initial flurry of adding books, I recorded little activity on the website until about 2014-2015 when I started using Goodreads in earnest. This graph doesn't really represent my reading history since there's a lot of missing data, but it does represent pretty well how I've used this website.

## Most popular author

I wondered which author I had read the most of. This is subject to recall bias, but it seems that #1 is Haruki Murakami, whose work I just love. Then came two comic artists (Gene Luen Yang and Sara Varon). Finally there were a whole bunch of 3-timers including Hugh Howey (Wool series), John Elder Robison (autistic writer), Stieg Larsson (The Girl Who... series) Suzanne Collins (Hunger Games series). 

```{r}
books %>% group_by(Author) %>% summarize(n = length(Author)) %>% arrange(desc(n))
```

Apparently I really like Haruki Murakami but everyone else is meh.

## My Ratings

I wondered about the ratings I had given books. 

```{r}
# Unrated books got a rating of zero
books$My.Rating <- ifelse(books$My.Rating == 0, NA, books$My.Rating)

books %>% filter(!is.na(My.Rating)) %>%
  ggplot(aes(x = My.Rating)) + geom_histogram(binwidth = 1) +
  ggtitle("My Ratings") +
  xlab("My Rating") + 
  theme_ipsum()
```

It looks pretty heavily skewed to 4 and 5 star ratings. In fact, my median rating was a 4.

```{r}
median(books$My.Rating, na.rm = T)
```

```{r}
books %>% 
  ggplot(aes(x=Average.Rating)) + geom_histogram(binwidth = 0.2) + 
  ggtitle("Community Ratings of Books that I've Added") +
  xlab("Rating") +
  theme_ipsum()
```

The median rating by the community was actually pretty similar to mine.

```{r}
median(books$Average.Rating)
```

## Difference between My Ratings and the Masses

Were there books that I enjoyed way more or less than the community? I didn't have the distribution of the ratings for each book, but I did have the mean and could calculate the difference between the community average rating and mine. 

```{r}
books %>% select(Title, Author, My.Rating, Average.Rating) %>% mutate(dRating = My.Rating - Average.Rating) %>% 
  filter(!is.na(dRating)) %>% arrange(desc(dRating)) %>%
  ggplot(aes(x = dRating)) +
  geom_histogram(binwidth = 0.25) +
  ggtitle("Difference between My Ratings and Community Ratings") +
  xlab("My Rating - Community Rating") +
  theme_ipsum()
```

Here are the top 10 books that I liked more than the community.
```{r}
books %>% select(Title, Author, My.Rating, Average.Rating) %>% mutate(dRating = My.Rating - Average.Rating) %>% 
  filter(!is.na(dRating)) %>% arrange(desc(dRating))
```

And the ones I liked worse than the community.
```{r}
books %>% select(Title, Author, My.Rating, Average.Rating) %>% mutate(dRating = My.Rating - Average.Rating) %>% 
  filter(!is.na(dRating)) %>% arrange(dRating)
```

Interestingly enough, John Elder Robison wrote #3 on the worse and #7 on the better lists.

## Publication Date

How old were the books I've been reading? 
```{r}
books %>% ggplot(aes(x = Original.Publication.Year)) + geom_histogram(binwidth = 5) +
  ggtitle("Publication Year of My Books") +
  xlab("Publication Year") + 
  theme_ipsum()
```

These two 17th century books were Shakespeare plays I had read before going to see them live (Othello, Twelfth Night). Taking those out led to this admittedly still skewed distribution.

```{r}
books %>% filter(Original.Publication.Year > 1700) %>%
  ggplot(aes(x = Original.Publication.Year)) + geom_histogram(binwidth = 2) +
  xlab("Original Publication Year") +
  ggtitle("Original Publication Year (Books Written since 1900)") +
  theme_ipsum()
```

The distribution is highly skewed, with a median original publication date of 2010.

```{r}
median(books$Original.Publication.Year, na.rm = T)
```

Those other outliers were:

```{r}
books %>% filter(Original.Publication.Year < 1980) %>% 
  arrange(Original.Publication.Year) %>% 
  select(Original.Publication.Year, Title) %>% 
  as.data.frame()
```

## Number of Pages (and over time)

The last thing I looked at was the number of pages I've read since beginning recording in earnest. 

```{r}
books %>% mutate(Year_read = factor(year(Date.Read))) %>%
  filter(!is.na(Year_read)) %>%
  group_by(Year_read) %>% 
  summarize(npages = sum(Number.of.Pages, na.rm = T)) %>%
  ggplot(aes(x = Year_read, y = npages)) +
  geom_bar(stat="identity") + 
  xlab("Year Read") + 
  ylab("Number of Pages") +
  ggtitle("Pages Recorded as Read per Year") +
  theme_ipsum()
```

This year (2017) has been a big reading year and it's not even half over yet. I think the summer reading program from my library and the Goodreads Reading Challenge have been big reasons that I have done so much this year.

## Date Published vs. Date Read

```{r}
books %>% select(Date.Read, Original.Publication.Year) %>% 
  filter(!is.na(Date.Read) & !is.na(Original.Publication.Year)) %>%
  ggplot(aes(x = Date.Read, y = Original.Publication.Year)) +
  geom_point() +
  theme_ipsum() +
  labs(title = "Date Read vs. Original Publication Year",
       x = "Date Read",
       y = "Original Publication Year")
```

Shoot, those Shakespeare plays really mess with the plot. I don't think even putting a log scale would help. I filtered them out to get the plot below.

```{r}
library(scales)
books %>% select(Date.Read, Original.Publication.Year) %>% 
  filter(!is.na(Date.Read) & !is.na(Original.Publication.Year)) %>%
  filter(Original.Publication.Year > 1700) %>% 
  ggplot(aes(x = Date.Read, y = Original.Publication.Year)) +
  geom_point() +
  theme_ipsum() +
  labs(title = "Date Read vs. Original Publication Year",
       x = "Date Read",
       y = "Original Publication Year")
```

